---
title: "Venables03"
author: "Robert A. Stevens"
date: "March 14, 2015"
output: html_document
---

# R Workshop Session 3:  Generalized Linear and Generalized Additive Models

Bill Venables, CSIRO, Australia

UseR! 2012, Nashville, 11 June, 2012

# Contents

1 An example from MASS: low birth weight  
1.1 Automated screening of variables  
1.2 An extended model with smooth terms  
1.3 Looking at the terms  
1.4 A helper function: the most frequent value  
1.5 The main two-way interaction  
2 Tiger prawn species split  
2.1 An initial GLM  
2.2 A long-term trend?  
2.3 A working GAM with new technology  
2.4 The spatio-temporal effect  
3 Technical highlights  
References  
Session information

# 1 An example from MASS: low birth weight

From Venables and Ripley (2002, Chap. 7).

- low:  indicator of birth weight less than 2.5 kg.

- age:  mother's age in years.

- lwt:  mother's weight in pounds at last menstrual period.

- race:  mother's race ('1' = white, '2' = black, '3' = other).

- smoke:  smoking status during pregnancy.

- ptl:  number of previous premature labours.

- ht:  history of hypertension.

- ui:  presence of uterine irritability.

- ftv:  number of physician visits during the first trimester.

- bwt:  birth weight in grams.

The original MASS code:

```{r, comment=NA}
library(MASS)
attach(birthwt)
race <- factor(race, labels = c("white", "black", "other"))
table(ptl)
ptd <- factor(ptl > 0)
table(ftv)
ftv <- factor(ftv)
levels(ftv)[-(1:2)] <- "2+"
table(ftv) # as a check
bwt <- data.frame(low = factor(low), 
                  age, 
                  lwt, 
                  race,
                  smoke = (smoke > 0), 
                  ptd, 
                  ht = (ht > 0), 
                  ui = (ui > 0), 
                  ftv)
detach() 
rm(race, ptd, ftv)
```

My preference now:

```{r, comment=NA}
suppressPackageStartupMessages(library(SOAR)) # picky!
suppressPackageStartupMessages(library(MASS))
BirthWt <- within(birthwt, {
  race <- factor(race, labels = c("white", "black", "other"))
  ptl <- ptl > 0
  ftv <- factor(ftv)
  levels(ftv)[-(1:2)] <- "2+"
  low <- factor(low, labels = c("normal", "low"))
  smoke <- (smoke > 0)
  ht <- (ht > 0)
  ui <- (ui > 0)
  bwt <- NULL ## remove actual birth weight
})
Store(BirthWt)
head(BirthWt, 2)
```

Advice from the fortunes package:

*If I were to be treated by a cure created by stepwise regression, I would prefer voodoo.* [Dieter Menne (in a thread about regressions with many variables) R-help (October 2009)]

- Automated screening is more defensible in cases of pure prediction.

- Automated screening is dangerous if used for inference.

Caveat emptor!

We reduce some of the clutter with:

```{r, comment=NA}
options(show.signif.stars = FALSE)
stepAIC <- function(..., trace = FALSE) # change default
  MASS::stepAIC(..., trace = trace)
dropterm <- function(..., sorted = TRUE) # change default
  MASS::dropterm(..., sorted = sorted)
```

## 1.1 Automated screening of variables

A starting point, main effects only:

```{r, comment=NA}
BW0 <- glm(low ~ ., binomial, BirthWt)
dropterm(BW0, test = "Chisq")
```

Screen for possible interactions:

```{r, comment=NA}
sBW0 <- stepAIC(BW0, scope = list(lower = ~1, upper = ~.^2))
dropterm(sBW0, test = "Chisq")
```

## 1.2 An extended model with smooth terms

We consider some flexibility in the age term and its interaction with ftv.

```{r, comment=NA}
suppressPackageStartupMessages(require(mgcv))
BW1 <- gam(low ~ smoke*ui + ht + s(lwt) + ptl + s(age) +
                 poly(age, 2)*ftv, 
           family = binomial, data = BirthWt)
anova(BW1)
```

## 1.3 Looking at the terms

First the "main effect" terms. This is a bit tricky...

```{r, comment=NA}
(nam <- names(model.frame(BW1)))
layout(matrix(1:8, 2, 4, byrow = TRUE))
termplot(BW1, terms = nam[2:7], se = TRUE) # fixed
plot(BW1) # smooth
```

The first nam is the response and the 8th and 9th refer to the smooth terms. termplot can handle non-smoothed terms, but smooth terms must be handled by the plot method for gam objects.

## 1.4 A helper function: the most frequent value

```{r, comment=NA}
mostFreq <- function(x, ...) UseMethod("mostFreq")
mostFreq.numeric <- stats::median.default  # check argument names
mostFreq.logical <- function(x, ...) {
  tx <- as.vector(table(x))
  tx[2] > tx[1]
}
mostFreq.character <- function(x, ...) {
  tx <- table(x)
  names(tx)[which.max(tx)]
}
mostFreq.factor <- function(x, ...)
  mostFreq.character(as.character(x))
Store(list = ls(pattern = "^mostFreq"))
```

## 1.5 The main two-way interaction

Predict the probability of low birth weight with varying age and ftv, and other variables at or near their modal value.

```{r, comment=NA}
all.vars(formula(BW1))
pBirthWt <- with(BirthWt,
                 expand.grid(smoke = mostFreq(smoke), 
                             ui = mostFreq(ui),
                             ht = mostFreq(ht), 
                             lwt = mostFreq(lwt),
                             ptl = mostFreq(ptl), 
                             age = min(age):max(age),
                             ftv = levels(ftv)))
pBirthWt$pBW1 <- predict(BW1, pBirthWt, type = "response")
library(lattice)
(ageFtv <- xyplot(pBW1 ~ age|ftv, pBirthWt, layout = c(3,1),
                  type = "l", ylab = "Pr(low birth wt)",
                  ylim = 0:1, aspect = 1))
```

To bring the predictions closer to the actual data, confine the predictions to age ranges that apply within the levels of ftv

```{r, comment=NA}
pBirthWt <- within(pBirthWt, {
  rngs <- do.call(cbind, with(BirthWt, tapply(age, ftv, range)))
  pBW1a <- pBW1
  is.na(pBW1a[age < rngs[1, ftv] | age > rngs[2, ftv]]) <- TRUE
  rm(rngs)
})
(ageFtvA <- xyplot(pBW1a ~ age|ftv, pBirthWt, layout = c(3,1),
                   type = "l", ylab = "Pr(low birth wt)",
                   ylim = 0:1, aspect = 1))
```

# 2 Tiger prawn species split

The Northern Prawn Fishery: Tiger prawn effort and "the blue line"

Background

- To species of Tiger prawns are caught together.

- Both species require separate Stock Assessment.

- The assessment model requires catches of Tiger prawns to be split (by weight)

- Problem: Build a model for partitioning catches into the two compoent species.

- Data: independent surveys (12 in all) where catches have been split into the two species, Penaeus semisulcatus (Grooved) and P. esculentus (Brown).

- Both species have annual offshore migration patterns.

Variables available:

- Response: Psem, Pesc, (Total = Psem + Pesc) in gms; Predictors:

- Longitude, Latitude – of trawl shot;

- Coast, Sea – alternative spatial coordinates;

- Depth – of trawl shot;

- Mud – the % mud in the substrate;

- DayOfYear – to allow for annual migration periodicity;

- ElapsedDays – days since 1970-01-01, for long term trend

- Survey – used for a random effect extension.

Strategy:

- Build a simpler GLM using mainly splines, with a term in DayOfYear and Sea to allow for temporal (annual migration) effects.

- Develop a more sophisticated GAM to take advantage of more recent modelling technology

- Look at a long-term trend term as a perturbation to the model

- Consider GLMMs with random terms for Survey, eventually

Model terms, GLM:

- Spline in Coast surrogate for large-scale benthic changes,

- Splines in Sea, Depth and Mud – more local spatial effects,

- Periodic term in DayOfYear and its interaction with Sea – annual migration effects,

- Spline in ElapsedDays – testing for long-term stability.

## 2.1 An initial GLM

The GLM fitting process is slow to converge under the normal algorithm. Two possible alternatives:

- Use the glm2 library, which has a modified convergence process, (Marschner, 2011).

- In this case the problem is with the variable weights needed. Fit a model ignoring weights, and use the linear predictor as a starting value for the weighted fit.

The periodic terms will use Fourier polynomials:

```{r, comment=NA}
Annual <- function (day, k = 4) { ## day of the year, starting from 0
  theta <- 2*base::pi*day/364.25
  X <- matrix(0, length(theta), 2 * k)
  nam <- as.vector(outer(c("c", "s"), 1:k, paste, sep = ""))
  dimnames(X) <- list(names(day), nam)
  m <- 0
  for (j in 1:k) {
    X[, (m <- m + 1)] <- cos(j * theta)
    X[, (m <- m + 1)] <- sin(j * theta)
  }
  X 
}

setwd("~/GitHub/Venables")
Tigers <- read.csv("Tigers.csv", header = TRUE, sep = ";")
library(splines)
temp <- glm(Psem/Total ~ ns(Coast, 10) + ns(Sea, 5) + ns(Depth, 5) +
              ns(Mud, 4) + Annual(DayOfYear, 4)*Sea, family = quasibinomial,
              data = Tigers, trace = TRUE)  # unweighted
Tigers$eta <- predict(temp)
TModelGLM <- update(temp, etastart = eta, weights = Total)
rm(temp)
Tigers$eta <- predict(TModelGLM)
TModelGLM$call$trace <- NULL  # for future updating
Store(TModelGLM)
(nam <- names(model.frame(TModelGLM))) # for term plotting
nam <- nam[2:7] # terms to plot
```

Look at the shape of the main effect terms, to see implications:

```{r, comment=NA}
layout(matrix(1:6, 2, 3, byrow = TRUE)) # 2 x 3 array of plots
termplot(TModelGLM, terms = nam, se = TRUE, rug = TRUE)
```

## 2.2 A long-term trend?

The stability of species ratios over time is important. We can check for this by including a spline term in ElapsedDays:

```{r, comment=NA}
TM2 <- update(TModelGLM, . ~ . + ns(ElapsedDays, 7))
anova(TModelGLM, TM2, test = "F")
```

Significant, but is it important?

```{r, comment=NA}
layout(matrix(1, 1, 1, byrow = TRUE))
termplot(TM2, terms = "ns(ElapsedDays, 7)", se = TRUE, rug = TRUE)
```

## 2.3 A working GAM with new technology

The mgcv package represents a major advance in smooth model fitting technology in sevaral respects, including

- Smoothed terms in multiple predictors can now be handled,

- A wide variety of basis functions is available, including e.g. thin plate splines, cyclic spline bases, etc.

- A powerful visualisation tool in projections of predictor variable space is avialable in addition to tools for inspection of individual terms

The price is:

- The package is still under development and new versions are fairly common (though becoming less so)

- The implementation is to some extent non-standard R. 

The working model:

```{r, comment=NA}
require(mgcv)
Attach()
TModelGAM <- gam(Psem/Total ~ s(Longitude, Latitude) +
                 te(DayOfYear, Sea, k = c(5, 5), bs = c("cc", "cs")) +
                 te(DayOfYear, Depth, k = c(5, 5), bs = c("cc", "cs")) +
                 te(Sea, Depth, k = c(5, 5), bs = "cs") +
                 s(Mud, k = 5),
                 family = quasibinomial, data = Tigers,
                 knots = list(DayOfYear = seq(0, 365.25, length=5)),
                 weights = Total, control = gam.control(trace = TRUE))
TModelGAM_NS <- update(TModelGAM, . ~ . + s(ElapsedDays))  # non-stationary
```

Some notes:

- An isotropic spatial term in Longitude and Latitude to account for purely spatial effects;

- Tensor spline (non-isotropic) terms in each pair of DayOfYear, Depth and Sea to account for temporal and environmental effects;

- A smooth term in Mud, also for environmental effects;

- The terms in DayOfYear are periodic with period 365 days (guaranteed by the data) in that coordinate bs = "cc";

- Other terms use a smooth spline basis, bs = "cs". Other choices of bases are available.

Some views of the fit:

```{r, comment=NA}
Oz <- read.csv("Oz.csv", header = TRUE, sep = ";")
par(las = 1)
layout(rbind(1:2))
plot(TModelGAM_NS, select = 1, asp = 1)
lines(Oz, col = grey(0.8))
title(main = "Lon x Lat, isotropic")
for(j in 2:6) plot(TModelGAM_NS, select = j)
vis.gam(TModelGAM_NS, view = c("Longitude","Latitude"))
title(main = "Lon x Lat, isotropic")
vis.gam(TModelGAM_NS, view = c("DayOfYear","Sea"))
title(main = "Day of year x Sea")
vis.gam(TModelGAM_NS, view = c("DayOfYear","Depth"))
title(main = "Day of year x Depth")
vis.gam(TModelGAM_NS, view = c("Depth","Mud"))
title(main = "Depth x Mud")
```

## 2.4 The spatio-temporal effect

Finally, we look at daily predictions for one year at 4 locations within the Gulf of Carpentaria:

Note that the migration effect is strongest in the Vanderlins islands region, where the Tiger prawn catch is high.

The prediction code is not shown, but the results are stored in a data frame called Data4 [not provided?]. The graphic is generaged by:

    Attach()
    require(lattice)
    print(xyplot(Fsemi ~ DayOfYear | Place, Data4, type = "l",
      ylab = expression(italic(P.)~~italic(semisulcatus)~~plain(proportion)), 
      xlab = "Day of the year", aspect = 0.7))

Note the device for mixed fonts in the annotations.

# Technical highlights

- Using within for neat data manipulation.

- Creating a temporary override function with changed defaults.

- stepAIC and dropterm.

- Suppressing package startup messages; simple gam. 

- termplot and finding the terms that can be plotted. 

- Elementary S3 generic and methods.

- all.vars finding all variables in an expression.

- The glm2 package for stable GLM fitting.

- Coddling a troublesome GLM fit with starting values.

- *et seq.*: Sophisticated GAM modelling and visualisation.

# References

Marschner, I. C. (2011, December). glm2: Fitting generalized linear models with convergence problems. *The* **R** *Journal* 3(2), 12–15.

Venables, W. N. and B. D. Ripley (2002). *Modern Applied Statistics with S* (Fourth ed.). New York: Springer. ISBN 0-387-95457-0.

# Session information

```{r, comment=NA}
sessionInfo()
```
