# R Workshop Session 1: Preliminaries
## Bill Venables, CSIRO, Australia
## UseR! 2012, Nashville, 11 June, 2012

# Contents

1 Some preliminaries: useful protocols and gratuitous advice

1.1 Use the file system

1.2 The SOAR package

1.3 Keep related objects together

1.4 Do not use attach()

1.5 Working protocols

2 Using R: some familiar concepts

References

Session information

# 1 Some preliminaries: useful protocols and gratuitous advice

## 1.1 Use the file system

In working with R, use the file system. A good protocol is:

* For each new project, set up a working directory which will contain all the files needed for that project in one place.

* The working directory may contain sub-directories for natural entities such as Data, Fig, Archive, Scripts, Code, &c

* It is better to start R in the working directory rather than start elsewhere use the GUI or setwd() to go there.

* Use the SOAR package (next slide) to keep objects available from one session to the next, and discard others. Keep it clean.

## 1.2 The SOAR package

* Use for keeping objects from one session to the next, especially useful for large objects, or objects requiring a lot of time to generate.

* Keeps .RData files of stores objects in a sub-directory of the working directory, ./.R_Cache.

* Store(...): place objects in cache, removing from memory, but still visible as promises,

* Objects() (or Ls()): list cache contents,

* Attach(): place the cache on the search path as promises

* Remove(...): delete objects from the cache, permanently.

An additional function, Search(), gives and enhanced view of the current search path.

## 1.3 Keep related objects together

### Bad:

```{r}
rm(list = ls())
library(SOAR)
set.seed(1234)
x <- sort(runif(500, 0, 10))
beta <- runif(5, -1, 1)
eta <- cbind(1, poly(x, 4)) %*% beta
y <- exp(eta + rnorm(500, 0, 0.1))
mu <- exp(eta + 0.1^2/2)
dummyData <- data.frame(x = x, mu = mu, y = y, eta = eta)
Store(dummyData)
ls()
Ls()
```

### Good:

```{r}
library(SOAR)
rm(list = ls())
set.seed(1234)
dummyData <- within(data.frame(x = sort(runif(500, 0, 10))), {
  beta <- runif(5, -1, 1)
  eta <- cbind(1, poly(x, 4)) %*% beta
  y <- exp(eta + rnorm(500, 0, 0.1))
  mu <- exp(eta + 0.1^2/2)
  rm(beta)
})
Store(dummyData)
ls()
head(dummyData, 2)
plot(y ~ x, dummyData, pch = 20, col = "red")
lines(mu ~ x, dummyData, col = "blue")
lines(exp(eta) ~ x, dummyData, col = "darkgreen", lty = "dashed")
```

## 1.4 Do not use attach()

* Visibility is key to ensuring R gets the right object.

### Bad:

```{r}
attach(dummyData)
beta <- coef(lm(eta ~ poly(x, 4))) ## OK so far
LModel <- lm(log(y) ~ poly(x, 4)) ## object incomplete
rbind(beta = beta, beta_hat = coef(LModel))
```

The object LModel relies on the context for its meaning. From where does the data come?

These are much worse, meaning unclear and prediction impossible.

```{r}
rough_1 <- lm(log(dummyData$y) ~ poly(dummyData$x, 4)) ## BAD!
rough_2 <- lm(log(dummyData[, 3]) ~ poly(dummyData[, 1], 4)) ## Horrible!
```

### Good:

```{r}
beta <- with(dummyData, qr.coef(qr(cbind(1, poly(x, 4))), eta))
LModel <- lm(log(y) ~ poly(x, 4), dummyData)
```

The object LModel now has information on where the data comes from.

Keep an eye on the search path as you work and keep it tidy:

```{r}
Search() ## From the SOAR package - enhanced
detach("dummyData")
```

## 1.5 Working protocols

* Find a front-end to R with which you feel comfortable. None is ideal (as yet). The following two are cross-platform.

1. Rstudio is probably best for beginner;

2. Emacs + ESS has a very steep learning curve;

* Establish your primary data sources early.

* Use scripts! This is very important.

* Do not use absolute file names in scripts! Your file names should be relative to the working directory.

* Establish, via scripts, a clear path from your primary data sources to R, and be prepared for changes.

* Use SOAR (or equivalent) to hold objects over temporarily from one session to the next, but do not rely on the saved object versions.

* Keep your global environment clean and your saved .RData file small. This will make startup quicker and keep your memory size in check.

A final look at the dummy example:

```{r}
ci <- confint(LModel)
data.frame(ci, beta = beta,
             OK = ifelse(ci[ , 1] < beta & beta < ci[ , 2], "yes", "no"),
             check.names = FALSE)
```

# 2 Using R: some familiar concepts

* R is a language for manipulating objects.

* In R, everything is an object and every object has a class.

* The R evaluator is (recursively) given an object manipulation task (function call) and names of objects to use.

* Where objects are located is governed by the scoping rules, which ultimately lead to the global environment and search path.

* Generic manipulations (print, plot, summary, &c) have their detailed operation determined by the class of the objects on which they act.

# References

Venables, W. N. and B. D. Ripley (2002). Modern Applied Statistics with S (Fourth ed.). New York: Springer. ISBN 0-387-95457-0.

# Session information

sessionInfo()
